---
title: "baconr: An rstan implementation of the Bayesian Age-Depth model *Bacon* (Blaauw and Christen, 2011)."
output: 
  html_document: 
    keep_md: yes
---

-------------------------------

**baconr** implements the Bayesian Age~Depth model, **Bacon**, in the **Stan** probabilistic programming language. It is at a very early stage of development and at the moment implements only the core non-Gaussian AR1 model described in Blaauw and Christen (2011). There is as yet no ability to convert between ^14^C ages and calibrated calendar ages.

There are currently just three exported functions: 

* `make_stan_dat` prepares data and parameter values into the correct format to be passed to the Stan model 
* `stan_bacon` calls a pre-compiled Stan implementation of Bacon and estimates the age model
* `plot_stan_bacon` plots a sample from the posterior distribution of the estimated age model

**baconr** also includes the example data, core MSB2K, included in the existing C++ implementation of [Bacon v2.2](http://www.chrono.qub.ac.uk/blaauw/bacon.html)

The motivation for creating this package is to make use of the Bacon age modelling routine more flexible, and to make further development 
of the age model itself easier, by coding it in a widely used higher-level probabilistic programming language.



*  Blaauw, Maarten, and J. Andrés Christen. 2011. Flexible Paleoclimate Age-Depth Models Using an Autoregressive Gamma Process. Bayesian Analysis 6 (3): 457-74. doi:10.1214/ba/1339616472.

*  Stan Development Team. 2016. Stan Modeling Language Users Guide and Reference Manual, Version 2.14.0.   http://mc-stan.org



## Installation

**baconr** can be installed directly from github

```{r, eval=FALSE}
if (!require("devtools")) {
  install.packages("devtools")
}

devtools::install_github("andrewdolman/baconr")
```


## Development

We aim to develop this package following the [Guidelines for R packages providing interfaces to Stan](https://cran.r-project.org/web/packages/rstantools/vignettes/developer-guidelines.html)


## Example


```{r, message = FALSE, warning=FALSE}
library(baconr)
library(knitr)
library(ggplot2)
library(tidyr)
library(dplyr)

opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE, cache = TRUE,
               fig.width = 6, fig.pos = "H", dpi = 150, autodep = TRUE)

```

### Data and parameters

the output from `make_stan_dat` is shown here to illustrate the required data and parameter format, but a separate call to `make_stan_dat` is not normally necessary as it is used internally by `stan_bacon`

```{r}
# Get number of sections K, so that they will be ~ 5cm
K_for_5cm <- round(diff(range(MSB2K$depth)) / 5)

make_stan_dat(depth = MSB2K$depth, 
  obs_age = MSB2K$age, 
  obs_err = MSB2K$error,
  K = K_for_5cm, nu = 6,
  acc_mean = 20, acc_var = "default",
  mem_mean = 0.7, mem_strength = 4)

```


### Fit the bacon model with `stan_bacon`

```{r}
fit <- stan_bacon(
  depth = MSB2K$depth, 
  obs_age = MSB2K$age, 
  obs_err = MSB2K$error,
  K = K_for_5cm, nu = 6,
  acc_mean = 20, acc_var = "default",
  mem_mean = 0.7, mem_strength = 4,
  iter = 2000, chains = 4)
```


```{r}
options(width = 85)
print(fit$fit, par = c("c_ages"))
```


### Plot the estimated age model

```{r}
set.seed(20170406)
plot_stan_bacon(fit, 100)
```


### Fit again at higher vertical resolution


```{r results='hide'}
fit2 <- stan_bacon(
  depth = MSB2K$depth, 
  obs_age = MSB2K$age, 
  obs_err = MSB2K$error,
  K = K_for_5cm*10, nu = 6,
  acc_mean = 20, acc_var = "default",
  mem_mean = 0.7, mem_strength = 4,
  iter = 2000, chains = 4)
```


```{r}
set.seed(20170406)
plot_stan_bacon(fit2, 100)
```

