---
title: "hamstr: Hierarchical Accumulation Modelling with Stan and R."
output: 
  html_document: 
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

-------------------------------

**hamstr** implements a the Bayesian Age~Depth model, **Bacon**, in the **Stan** probabilistic programming language. It is at a very early stage of development and at the moment implements only the core non-Gaussian AR1 model described in Blaauw and Christen (2011). Functions from the R package **Bchron** (Parnell 2016) can be used to calibrate ^14^C ages to calendar ages.


There are currently just three exported functions: 

* `make_stan_dat_hamstr` prepares data and parameter values into the correct format to be passed to the Stan model 
* `stan_hamstr` calls a pre-compiled Stan implementation of Bacon and estimates the age model
* `plot_stan_hamstr` plots a sample from the posterior distribution of the estimated age model

**hamstr** also includes the example data, core MSB2K, included in the existing C++ implementation of [Bacon v2.2](http://www.chrono.qub.ac.uk/blaauw/bacon.html)

The motivation for creating this package is to make use of the Bacon age modelling routine more flexible, and to make further development 
of the age model itself easier, by coding it in a widely used higher-level probabilistic programming language.


*  Parnell, Andrew. 2016. Bchron: Radiocarbon Dating, Age-Depth Modelling, Relative Sea Level Rate Estimation, and Non-Parametric Phase Modelling. R package version 4.2.6. https://CRAN.R-project.org/package=Bchron

*  Blaauw, Maarten, and J. Andrés Christen. 2011. Flexible Paleoclimate Age-Depth Models Using an Autoregressive Gamma Process. Bayesian Analysis 6 (3): 457-74. doi:10.1214/ba/1339616472.

*  Stan Development Team. 2016. Stan Modeling Language Users Guide and Reference Manual, Version 2.14.0.   http://mc-stan.org




## Installation

**hamstr** can be installed directly from Github

```{r, eval=FALSE}
if (!require("devtools")) {
  install.packages("devtools")
}

devtools::install_github("andrewdolman/hamstr", args = "--preclean", build_vignettes = FALSE)
```


## Development

We aim to develop this package following the [Guidelines for R packages providing interfaces to Stan](https://cran.r-project.org/web/packages/rstantools/vignettes/developer-guidelines.html)


## Example


```{r, message = FALSE, warning=FALSE}
library(hamstr)
library(knitr)
library(ggplot2)
library(tidyr)
library(dplyr)
library(Bchron)

opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE, cache = TRUE,
               fig.width = 6, fig.pos = "H", dpi = 300, autodep = TRUE)

```

### Data and parameters

First convert 14^C ages to calendar ages. Functions from the R package **Bchron** can be used for this.

```{r}
cal.ages <- Bchron::BchronCalibrate(ages=MSB2K$age,
                            ageSds=MSB2K$error,
                            calCurves=rep("intcal13", nrow(MSB2K)),
                            ids=paste0("Date-", 1:nrow(MSB2K)))

SimplifyCalAge <- function(ages, densities){
  
  wts <- densities / sum(densities)
  
  M <- sum(wts > 0)

  w.mean <- sum(ages * wts)
  w.std <- sqrt(sum(wts * (ages-w.mean)^2) / ((M-1)/M * sum(wts)))
  
  return(c("mean" = w.mean, "std" = w.std))
  
}


MSB2K$age.cal <- sapply(cal.ages, function(x) SimplifyCalAge(x$ageGrid, x$densities)["mean"])
MSB2K$age.cal.sd <- sapply(cal.ages, function(x) SimplifyCalAge(x$ageGrid, x$densities)["std"])
```


the output from `make_stan_dat_hamstr` is shown here to illustrate the required data and parameter format, but a separate call to `make_stan_dat_hamstr` is not normally necessary as it is used internally by `stan_hamstr`

```{r}
# Get number of sections K, so that they will be ~ 5cm
K_for_5cm <- round(diff(range(MSB2K$depth)) / 5)

stan_dat <- make_stan_dat_hamstr(depth = MSB2K$depth, 
  obs_age = MSB2K$age.cal, 
  obs_err = MSB2K$age.cal.sd,
  K = K_for_5cm, nu = 6,
  acc_mean = 20, acc_alpha = 1.5,
  mem_mean = 0.7, mem_strength = 4)

stan_dat
```

```{r, echo=FALSE}
# Examine scaling behaviour of memory prior
# Behaves very strangely when deltaC is small and mem.mean is small

# hamstr does not match output of Bacon2.2 when mem_mean = 0.1, mem_strength = 4
# in hamstr the prior is strong and dominates, in Bacon2.2 the data dominate the prior

#(0.1^0.1)^10

# priorwU <- function(w, deltaC, alpha, beta) {
# 			    ds=1.0
# 			    logw=log(w)
# 			    tmp <- (ds/deltaC)*(1.0-alpha)*logw + (1.0-beta)*log(1.0-exp((ds/deltaC)*logw))
# 			    tmp #/ sum(tmp, na.rm = T)
# 			}
# x <- seq(0, 1, length.out = 1000)
# 
# plot(x, dbeta(x, stan_dat$mem_alpha, stan_dat$mem_beta), type = "l")
# plot(x, exp(-priorwU(w = x, deltaC = 1, alpha = stan_dat$mem_alpha, beta = stan_dat$mem_beta))
#      , type = "l", col = "Green")
# plot(x, exp(-priorwU(w = x, deltaC = 5, alpha = stan_dat$mem_alpha, beta = stan_dat$mem_beta))
#      , type = "l", col = "Red")
``` 


### Fit the bacon model with `stan_hamstr`

```{r}
fit <- stan_hamstr(
  depth = MSB2K$depth, 
  obs_age = MSB2K$age.cal, 
  obs_err = MSB2K$age.cal.sd,
  K = K_for_5cm, nu = 6,
  acc_mean = 20, acc_alpha = 1.5,
  mem_mean = 0.7, mem_strength = 4,
  iter = 2000, chains = 3)
```


```{r}
options(width = 85)
print(fit$fit, par = c("R", "w", "c_ages"))
```


### Plot the estimated age model

```{r bacon_defaults}
set.seed(20170406)
plot_stan_hamstr(fit, 1000)
```


### Fit again with stronger prior on accumulation rates


```{r, results='hide'}
fit2 <- stan_hamstr(
  depth = MSB2K$depth, 
  obs_age = MSB2K$age.cal, 
  obs_err = MSB2K$age.cal.sd,
  K = K_for_5cm, nu = 6,
  acc_mean = 20, acc_alpha = 25,
  mem_mean = 0.7, mem_strength = 4,
  iter = 2000, chains = 3)
```


```{r stronger_prior}
plot_stan_hamstr(fit2, 1000)
```


### Fit again with strong prior for lower memory


```{r, results='hide'}
fit3 <- stan_hamstr(
  depth = MSB2K$depth, 
  obs_age = MSB2K$age.cal, 
  obs_err = MSB2K$age.cal.sd,
  K = K_for_5cm, nu = 6,
  acc_mean = 20, acc_alpha = 1.5,
  mem_mean = 0.1, mem_strength = 4,
  iter = 2000, chains = 3)
```


```{r weaker_prior}
plot_stan_hamstr(fit3, 1000)
```


### Add tephras at 25 and 90 cm


```{r, results='hide'}
teph <- data.frame(depth = c(25, 87.5), age.cal = c(5000, 6500), age.cal.sd = c(3, 5))
MSB2K.2 <- bind_rows(MSB2K, teph) %>% 
  arrange(age.cal)
  

fit4 <- stan_hamstr(
  depth = MSB2K.2$depth, 
  obs_age = MSB2K.2$age.cal, 
  obs_err = MSB2K.2$age.cal.sd,
  K = K_for_5cm, nu = 6,
  acc_mean = 20, acc_alpha = 1.5,
  mem_mean = 0.7, mem_strength = 4,
  iter = 2000, chains = 3)
```


```{r tephras}
plot_stan_hamstr(fit4, 1000) +
  geom_pointrange(data = teph,
                  aes(x = depth, y = age.cal, ymax = age.cal + age.cal.sd, ymin = age.cal - age.cal.sd),
                  group = NA, colour = "Blue", alpha = 0.5) 
```


### Add hiatus

```{r fit_hiatus, results="hide"}
# Add 1000 years to all depths below 40 cm
hi_ages <- MSB2K$age
hi_ages[MSB2K$depth > 40] <- hi_ages[MSB2K$depth > 40] + 1000

fit_hi <- stan_hamstr(
  depth = MSB2K$depth, 
  obs_age = hi_ages, 
  obs_err = MSB2K$age.cal.sd,
  hiatus_depth = c(40),
  hiatus_length = c(1000),
  K = 20, nu = 6,
  acc_mean = 20, acc_alpha = 1.5,
  mem_mean = 0.7, mem_strength = 4,
  iter = 2000, chains = 3)
```

```{r hiatus}
plot_stan_hamstr(fit_hi)
```



### Distribution of sediment accumulation rates

```{r}
age.fit <- fit
age.mod <- rstan::extract(age.fit$fit)

bp.dat <- age.mod$x[1:3000,] %>% 
  as_tibble() %>% 
  tibble::rownames_to_column("Rep") %>% 
  gather(Depth, value, -Rep) %>% 
  mutate(Depth = as.numeric(gsub("V", "", Depth)),
         Depth = Depth * age.fit$data$delta_c,
         Rep = as.numeric(Rep))

# bp.dat.bw <- bp.dat %>% 
#   group_by(Depth) %>% 
#   summarise(mean.val = mean(value),
#             upr = quantile(value, 0.9),
#             lwr = quantile(value, 0.1))
# 
# bp.dat.bw %>% 
#   ggplot(aes(x = Depth)) + 
#   geom_pointrange(aes(y = mean.val, ymax = upr, ymin = lwr))
```

```{r acc_rates}
bp.dat %>% 
  ggplot(aes(x = Depth, y = 1/value, group = Depth)) + 
  geom_violin() +
  #geom_boxplot() +
  #geom_point(alpha = 0.15)
  scale_x_continuous("Depth [cm]")+
  scale_y_continuous("Sediment accumulation rate [cm/yr]",
                     trans = "log10", breaks = c(0.01, 0.05, 0.1, 0.5)) +
  annotation_logticks(sides = "l") +
  expand_limits(y = 0.01) + 
  theme_bw()
```

```{r acc_intervals}
bp.dat %>% 
  ggplot(aes(x = Depth, y = value, group = Depth)) + 
  geom_violin() +
  #geom_boxplot() +
  #geom_point(alpha = 0.15)
  scale_x_continuous("Depth [cm]")+
  scale_y_continuous("Sedimentation interval [yr/cm]",
                     trans = "log10", breaks = c(1, 5, 10, 20, 50)) +
  annotation_logticks(sides = "l") +
#  expand_limits(y = 0.01) + 
  theme_bw()
```



```{r, echo=FALSE}
# bp.dat %>% 
#   filter(Rep %in% sample(Rep, 110)) %>%  
#   ggplot(aes(x = Depth, y = 1/value, group = Rep, colour = as.factor(Rep))) + 
#   geom_line(alpha = 0.51) +
#   scale_y_continuous(trans = "log10", breaks = c(0.01, 0.05, 0.1, 0.5)) +
#   expand_limits(y = 0.01)

# plot(density(age.mod$alpha[,]), xlim = c(0, 50))
# lines(seq(0.1, 50, length.out = 100), dgamma(seq(0.1, 50, length.out = 100),
#                     shape = age.fit$data$acc_mean^2 / age.fit$data$acc_var,
#                     rate = age.fit$data$acc_mean/age.fit$data$acc_var),
#       col = "Red")
```


